---
title: "More seeds yield more shoots, but not higher per-seed establishment in a clam-mediated seagrass restoration experiment"
author: 
  - name: Cori Lopazanski
    affiliations:
      - id: duml
        name: Duke University
        department: Duke University Marine Lab, Nicholas School of the ENvironment
        address: 135 Duke Marine Lab Rd.
        city: Beaufort
        region: NC
        postal-code: 28516
    attributes:
      corresponding: true
  - name: Christina Cutshaw
    affiliations:
      - ref: duml
  - name: Mason Ibrahim
    affiliations:
      - ref: duml
  - name: Brian Silliman
    affiliations:
      - ref: duml
format: 
  pdf:
    mainfont: "Times New Roman"
editor_options:
  markdown:
    wrap: 80
execute: 
  echo: false
  warning: false
  error: false
---

## Question

What are the relative effects of seagrass seed density on clams (number of
seagrass seeds) and clam burial method on seagrass germination and growth?

## Hypotheses

-   Germination probability increases with seed density up to an intermediate
    optimum then levels off or declines due to space/competition

-   Manually buried clams produce higher germination and faster early growth
    than clams placed on the surface because manual burial reduces seed exposure
    and displacement

    -   *Applied angle:* Manual burial is more time-consuming and may limit
        scalability compared to scattering clams on the surface. If there is no
        difference in germination or growth with clam placement, then that would
        be a positive effect.

-   The benefit of burial is greater at low/moderate seed densities, at very
    high seed densities negative density effects reduce the advantage of burial
    (maybe – this element somewhat less important, but good to keep in mind
    given crossed factor)

## Methods

*Seed retrieval and storage*

*Zostera marina* seagrass seeds were retrieved by collecting reproductive shoots
from seagrass beds in Back Sound, NC in May 2025. Shoots were placed in
flow-through seawater tanks indoors in the lab and seeds were allowed to fall
naturally to the bottom of the tank. Fallen seeds were collected from the bottom
of the tank and placed in mesh bags in a flow-through tank. While this tank was
exposed to general ambient light conditions, the mesh bags reduced overall light
exposure. Before deployment, 100 seeds were randomly selected to test for
viability using tetrazolium staining \[add staining method\]. Viability: 89/100.

*Planting experiment*

We conducted a field experiment near Harker’s Island, North Carolina, to test
how seed density and clam burial behavior influences seagrass seed germination
and seagrass growth. Seagrass seeds were glued to live clams (\~2 cm diameter)
using a gelatin-based bioglue at one of ten density levels, spanning 10-100
seeds per clam in increments of 10 seeds. Seed density was crossed with two clam
placement treatments (manually buried or placed on sediment surface) in a full
factorial design with four replicates per treatment (80 plots total).

We used a gelatin-based bioglue to attach seeds to clams. The bioglue was
prepared by dissolving dry gelatin powder in saltwater (35 ppt) to a final
concentration of 9%. The mixture was heated to 35°C until fully dissolved and
then cooled to 28°C before application. A thin layer of bioglue was pipetted
onto one shell from each clam and allowed to partially set before seeds were
arranged in a single layer and covered with a second thin coat of glue. Clams
were refrigerated at 9°C for 30 minutes to solidify the glue, then transported
to the field in coolers and deployed the same day. The total time from initial
glue preparation through field deployment was less than four hours.

Clams with glued seeds were assigned to one of two burial treatments: manually
buried to 2 cm depth or placed on the sediment surface to allow self-burial.
Each clam was positioned at the center of a 0.25m^2^ plot, with the corner of
each plot marked with a lawn staple (10-inch length) to facilitate resampling.
Plots were spaced 0.5 m apart in a grid located on a sandy area with average
depth \~0.8 m and at least 10-m away from adjacent salt marsh, oyster reef, and
seagrass habitat.

Germination and seedling growth were monitored over time to quantify the effects
of seed density and burial treatment on seagrass establishment success. We
defined germination as the emergence of the first green shoot above the sediment
surface. Plots were monitored biweekly. For each plot during each survey, we
recorded the number of live shoots and the height of up to five random shoots
(nearest mm). We monitored clam presence using brief visual and tactile
confirmation near the original placement point, but avoided excavating clams to
limit seedling disturbance. We recorded evidence of clam predation (e.g., shell
fragments, empty shells), and whether the clam was live, dead, or not located.

Clam predation was extremely high. Within the first few days of the experiment,
the number of clams remaining was significantly reduced \[see data\]. By
mid-December, there were few clams remaining and therefore we stopped checking
for clams. We suspect blue crabs or bullnose rays to be primary predators.

-   Consider taking downward-facing photographs of each quadrat to quantify
    percent cover and spatial spread over time? Or at set intervals? Or at end?

-   Consider whether to destructively sample or leave to see if anything happens
    next year?

    At the end of the experiment, we collected a single sediment core (X
    diameter, Y depth) centered on the plot, encompassing both above- and
    belowground components. Aboveground shoots were clipped at the sediment
    surface, counted, and measured for length (nearest mm). Sediment cores were
    washed over a X-mm sieve to recover belowground biomass. Above- and
    belowground materials were dried at X C for X hours to constant mass.
    Remaining clams were excavated from each plot and recorded as live or dead;
    unrecovered clams were classified as lost (presumed displaced or predated).

-   Consider collecting sediment samples for organic matter or grain size?

*Data analyses*

This experiment is a fully factorial manipulation of seed density and clam
placement, with repeated observations of seagrass establishment (counts of live
shoots) and seagrass growth (shoot height) through time at the plot level. Seed
density (10-100 seeds per clam) was treated as a continuous predictor, and clam
placement was modeled as a factor (manually buried vs. placed on the sediment
surface).

```{r}
# Setup -------------------------------------------------------------------------

library(readxl)
library(tidyverse)
library(lme4)
library(glmmTMB)
library(DHARMa)

# Read data from Excel file 
dat <- read_xlsx("data/seed-density.xlsx", na = "NA") 

# Build ------------------------------------------------------------------------
# - density is continuous
# - burial is a factor
# - plot_id identifies the repeated-measures unit
# - time is measured as days since the first sampling date
# - shoot_count is the number of shoots observed at each sampling event

dat <- dat %>% 
  mutate(plot_id = factor(plot_id),
         burial = factor(burial_treatment,
                         levels = c(0, 1),
                         labels = c("Surface", "Buried")),
         date = lubridate::ymd(date),
         time_days = as.numeric(date - date('2025-10-31'))) %>% 
  # monitoring on 4 Jan 2026 was incomplete because poor conditions (remove)
  filter(date != '2026-01-04') 
  

```

We conducted initial exploratory analyses to characterize temporal patterns in
shoot density across seed density and clam placement treatments. First, we
visualized plot-level trajectories of shoot densities through time, stratified
by seed density to assess whether temporal dynamics differed across density
treatments and whether patterns varied with clam placement.

To highlight treatment-level patterns, we calculated average shoot density
across replicates for each sampling date, seed density, and clam placement
treatment (n = 4 plots for each combination). We plotted the temporal
trajectories of these treatment-level means to examine general trends in shoot
density over time and assess potential differences between clam placement
treatments. These descriptive analyses were used to identify broad patterns (or
lack thereof) and inform model structure.

As shoot density is inherently constrained by the number of seeds planted, we
also examined shoot density relative to planted seed density. Probability of
shoot establishment was quantified as the number of live shoots relative to the
number of seeds initially planted in each plot.

```{r}
# Calculate treatment-level means through time
dat_means <- dat %>% 
  group_by(date, density, burial) %>% 
  summarize(mean_shoot_count = mean(shoot_count))

# Calculate proportion established relative to seed density
dat_prob <- dat %>% 
  mutate(est_prob = shoot_count/density) %>% 
  select(plot_id, burial, density, time_days, shoot_count, est_prob) %>% 
  mutate(failures = density - shoot_count) %>% 
  mutate(density_c = density - mean(density, na.rm = TRUE))


```

[Modeling shoot establishment]{.underline} (in progress)

We modeled this response using generalized linear mixed models with a binomial
distribution, specifying the number of observed shoots as successes and the
initial seed density as the denominator. Models included fixed effects for
density, burial treatment, time (days since deployment), an interaction between
burial and density, and plot as a random intercept to account for repeated
sampling.

We'll do this a) across time, b) endpoint only, and c) with and without
quadratic terms.

Test for overdispersion and if present try betabinomial.

We're calling it "shoot establishment" instead of germination because we are
observing a cumulative net establishment (seed retention, germination, shoot
survival) at each sampling interval.

Start with a binomial mixed model where the response is the number of shoots
relative to the initial seeds. Each plot started with a set number of seeds, and
each sampling event we see a number of shoots. As a binomial process, the
"successes" are the seeds that produce shoots and the "failures" are seeds that
didn't. Answers the question "out of \# seeds, how many became shoots?" It also
accounts for the different starting number of seeds (different number of
'chances') and explicitly models the probability of establishment per seed,
answering questions about density-dependent net shoot establishment.

```{r}
# Build "shoot establishment" models 
# Binomial mixed model:
# - Response: shoots relative to initial seeds
# - Fixed: seed density, burial treatment, time
# - Random: plot_id
# Modeling germination/establishment as a binomial process - each plot started with a set # of seeds, and at each sampling time we see a number of shoots.
# The "success" = seeds that produced shoots, and "failures" = seeds that didn't.
# --- out of X seeds, how many became shoots?
# --- accounts for different starting number of seeds
# --- models the probability of germination/establishment per seed, not just raw
# --- answers q's about density-dependent net shoot establishment

mbinom <- glmmTMB(cbind(shoot_count, failures) ~ 
                    density + burial + time_days + density:burial + 
                    (1|plot_id), 
                  family = binomial, data = dat_prob)

summary(mbinom)

# Time is highly significant, positive = establishment increases over time
# Density terms are non-significant, so no evidence for increasing establishment with density or an intermediate optimum with density. No detectable density-dependent germination response across 10-100 seeds per clam. 
# Burial treatment not significant, so no overall difference between buried vs. surface treatments.
# Densiy x Burial interaction terms not significant = no evidence that burial effects depend on density. No support for context-dependent burial effects across densities. 
# Plot-level heterogeneity a major driver of outcomes.
# That combination suggests either density range is not limiting (10–100 all sufficient), or other processes dominate (e.g., disturbance, predation, microsite conditions)

res_mbinom <- DHARMa::simulateResiduals(mbinom)
plot(res_mbinom)
testDispersion(res_mbinom)


# Overdispersion is close - not significant but there... could use a betabinomial distribution instead.
mbetabinom <- glmmTMB(cbind(shoot_count, failures) ~ 
                    density + burial + time_days + density:burial + 
                    (1|plot_id), 
                  family = betabinomial(), data = dat_prob)
summary(mbetabinom)
res_mbetabinom <- DHARMa::simulateResiduals(mbetabinom)
plot(res_mbetabinom)
testDispersion(res_mbetabinom) #no better..

# Test a second model that uses centered density, so the effect sizes are relative to the mean density... 
mbinom2 <- glmmTMB(
  cbind(shoot_count, density - shoot_count) ~
    density_c +
    I(density_c^2) +
    burial +
    time_days +
    density_c:burial +
    I(density_c^2):burial +
    (1 | plot_id),
  family = betabinomial(),
  data = dat_prob
)




```

## Results

*Descriptive visualizations*

[Trajectories of raw shoot densities]{.underline}

Initial visualizations show that shoot densities increase over time, and that
higher seed densities generally yield higher shoot densities (Figure 1). This
pattern does not appear to vary depending on whether clams were manually buried
or placed on the sediment surface. For visualization purposes, these trends are
shown for each plot through time (Figure 1) as treatment-level means through
time (Figure 2), and as treatment-level means through time but grouped by clam
placement (Figure 3).

```{r plot-level-trajectories}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of shoot density (count per 0.0625 m²) from November 2025 through April 2026 across experimental plots, shown separately for initial seed densities from 10 to 100 seeds (facets). Each line represents a single plot. Clam placement treatments are indicated by symbol and line type: clams placed on the sediment surface are shown as green circles with solid lines, and manually buried clams as purple triangles with dashed lines."

ggplot(dat, aes(x = date, y = shoot_count, group = plot_id, color = burial)) +
  geom_line(aes(linetype = burial), alpha = 0.7) +
  geom_point(aes(shape = burial), alpha = 0.7) +
  facet_wrap(~density, nrow = 2) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       shape = "Clam placement",
       linetype = "Clam placement",
       color = "Clam placement") +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r treatment-level-trajectories}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of mean shoot density (count per 0.0625 m²) from November 2025 through April 2026, shown separately for initial seeding densities from 10 to 100 seeds (facets). Values represent treatment-level means (n = 4 plots per density × clam placement combination). Clam placement treatments are indicated by symbol and line type: clams placed on the sediment surface are shown as green circles with solid lines, and manually buried clams as purple triangles with dashed lines."

ggplot(dat_means, aes(x = date, y = mean_shoot_count, group = burial, color = burial)) +
  geom_line(aes(linetype = burial)) +
  geom_point(aes(shape = burial)) +
  facet_wrap(~density, nrow = 2) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       shape = "Clam placement",
       linetype = "Clam placement",
       color = "Clam placement") +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r treatment-trajectories-by-placement}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of mean shoot density (count per 0.0625 m²) from November 2025 through April 2026, shown separately by clam placement treatment (facets). Values represent treatment-level means (n = 4 plots per density × clam placement combination). Colors denote initial seed density from 10 to 100 seeds, with darker shades indicating higher densities."

ggplot(dat_means %>% 
         mutate(density = factor(density)),
       aes(x = date, y = mean_shoot_count, group = density, color = density)) +
  geom_line() +
  geom_point() +
  facet_wrap(~burial, nrow = 1) +
  scale_color_manual(values = c('#d9f0ea', '#b2e2d8', '#80cdc1', '#4db6ac', '#2ca25f',
                                '#238b45','#006d2c', '#005824','#00441b','#002e12')) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       color = "Seed density") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

[Proportion of established shoots]{.underline}

While shoot density increased with seed density, the *proportion* of seeds that
produced shoots did not. Here, the probability of shoot establishment is shown
based on the most recent sampling date: `r format(max(dat$date), "%B %d, %Y")`.

```{r}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Relationship between initial seed density and shoot establishment at the final sampling point. Establishment is defined as the proportion of seeds that resulted in observable shoots, reflecting the combined effects of seed retention, germination, and early seedling survival. Points represent individual plots; lines show simple linear model predictions (±95% CI) by clam placement treatment, on the sediment surface (solid circles) or manually buried (dashed triangles)."


ggplot(dat_prob %>% filter(time_days == 152), aes(x = density, y = est_prob, shape = burial, color = burial)) +
  geom_smooth(aes(x = density, y = est_prob, linetype = burial, color = burial, fill = burial),
              method = 'lm', alpha = 0.3) +
  geom_point(position = position_jitter(width = 0.6, height = 0.005), 
             alpha = 0.6, size = 3)+
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 6) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  scale_fill_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = "Seed density",
       y = "Proportion of seeds producing shoots",
       linetype = "Clam placement",
       color = "Clam placement",
       fill = "Clam placement",
       pch = "Clam placement") +
  theme_classic() +
  theme(legend.position = c("inside"),
        legend.position.inside = c(0.9, 0.8))

```

Figure 3 — Shoot growth (length) over time

Growth response.

X-axis: time Y-axis: shoot length Lines: predicted growth trajectories Facet by
density (e.g., 3–4 representative levels) Color by burial

Purpose:

Tests whether treatments affect early growth rather than establishment Shows
whether burial influences growth rate (even if germination doesn’t) Figure 4 —
Early predation effects on establishment

Mechanistic figure (important).

X-axis: seed density Y-axis: establishment probability Panels (facet): no early
predation early predation Lines: predicted relationships (with CI) Color: burial
(optional)

Alternative simpler version:

Boxplots or points of final establishment by early predation (yes/no)

Purpose:

Tests whether early disturbance suppresses establishment Provides explanation
for null treatment effects Links field observation to outcome

This is likely your most important mechanistic figure.

Figure 5 — Clam fate / predation over time (optional but strong)

Supports mechanism.

Option A (preferred):

Y-axis: probability of shell fragments (or live clam) X-axis: time Lines:
predicted probability Color: burial

Option B:

proportion of plots with shell fragments over time

Purpose:

Shows when predation occurs (early pulse) Tests whether burial reduces predation
risk Connects directly to

Figure 4 Recommended minimal set (if limiting to 3–4 figures)

If you want a tight paper:

Germination vs density (Fig 1) Germination over time (Fig 2) Early predation
effect (Fig 4) Growth trajectories (optional, if growth is meaningful)

Clam fate can move to supplement unless it is a central result.

Overall narrative structure across figures Fig 1: No strong density or burial
effect Fig 2: Time dominates establishment Fig 4: Early predation explains
variation Fig 3 (optional): Growth also largely insensitive

This sequence makes the paper read as:

“We expected density and burial to matter; they did not strongly. Instead, early
disturbance and temporal dynamics structure establishment.”

## **Notes**

[*Deployed Friday October 31, 2025 at 12pm (during low tide).*]{.underline}

We marked the corners of each plot with tall bamboo poles (3 m) hammered into
the sediment using a rubber mallet while standing on the stern of the boat. We
established transects for each row in the grid, marking the start and end of
each transect with small bamboo poles (1 m) sunk into the sediment (left
permanently during the experiment to facilitate resampling). Rows were oriented
roughly parallel to the adjacent marsh shoreline, with seed density treatments
increasing each row from the side of the site closest to the marsh edge (lowest
density) to the outermost row (highest density). Note that the site itself is
\~50m from the marsh edge. Facing the marsh, the four columns on the left side
of the site contained manually buried clams, and the four columns on the right
side of the site contained clams placed on the sediment surface (see site
diagram, below).

On at least two occasions, the bioglue detached from the clam during underwater
handling, though the seeds remained intact within the glue matrix. These cases
were treated as valid experimental conditions rather than handling errors, and
both the detached bioglue-seed cluster and the corresponding clam were placed in
the assigned treatment position. Similar detachment events were anecdotally
observed during other planting experiments that season. 

Conditions during deployment were fairly windy (15-25 mph) with air temperatures
18.3°C and water temperatures 21.7°C. Visibility was poor; less than 0.5 m,
which prevented any visual confirmation of activity of “surface” clams (e.g.,
movement, burial). Deployment was \~30 minutes after low tide (+0.47ft) and
water depth was \~0.5-0.8 m during deployment (rising; knee to hip). Sediment
was sand covered by a thin layer of mud with minimal detritus, though some
sparse *Halodule wrightii* was observed at the margins of the site, especially
towards the outermost row (highest density treatments). Upon re-examining the
site under better conditions, a few plots contained sparse *Halodule* within the
plot.

[*Site diagram (for internal purposes)*]{.underline}

*(All plots 0.25m2 and are equally spaced by 0.5m2)*

Letters mark tall bamboo poles, and x's mark short bamboo poles used to lay out
rows. Pole B is closest to Harker’s Island Boat Ramp. A-B parallel to marsh, B-C
closest to bridge and parallel to oysters, A-D farthest to bridge parallel
opposite oysters, D-C closest to entrance/exit to study area and adjacent
seagrass beds.

\~\~\~\~ MARSH EDGE, Mason + Christy Experiments \~\~\~\~

| A     |      |      |      |      |      |      |      |      | B     |
|-------|------|------|------|------|------|------|------|------|-------|
| x     | B10  | B10  | B10  | B10  | S10  | S10  | S10  | S10  | x     |
| x     | B20  | B20  | B20  | B20  | S20  | S20  | S20  | S20  | x     |
| x     | B30  | B20  | B20  | B20  | S30  | S30  | S30  | S30  | x     |
| x     | B40  | B40  | B40  | B40  | S40  | S40  | S40  | S40  | x     |
| x     | B50  | B50  | B50  | B50  | S50  | S50  | S50  | S50  | x     |
| x     | B60  | B60  | B60  | B60  | S60  | S60  | S60  | S60  | x     |
| x     | B70  | B70  | B70  | B70  | S70  | S70  | S70  | S70  | x     |
| x     | B80  | B80  | B80  | B80  | S80  | S80  | S80  | S80  | x     |
| x     | B90  | B90  | B90  | B90  | S90  | S90  | S90  | S90  | x     |
| x     | B100 | B100 | B100 | B100 | S100 | S100 | S100 | S100 | x     |
| **D** |      |      |      |      |      |      |      |      | **C** |

\~\~\~\~ ACCESS EDGE, towards deeper water \~\~\~\~
