---
title: "More seeds yield more shoots, but not higher per-seed establishment in a clam-mediated seagrass restoration experiment"
author: 
  - name: Cori Lopazanski
    affiliations:
      - id: duml
        name: Duke University
        department: Duke University Marine Lab, Nicholas School of the Environment
        address: 135 Duke Marine Lab Rd.
        city: Beaufort
        region: NC
        postal-code: 28516
    attributes:
      corresponding: true
  - name: Christina Cutshaw
    affiliations:
      - ref: duml
  - name: Mason Ibrahim
    affiliations:
      - ref: duml
  - name: Joseph P. Morton
    affiliations:
      - ref: duml
  - name: Brian Silliman
    affiliations:
      - ref: duml
format: 
  pdf:
    mainfont: "Times New Roman"
  docx: 
    fig-cap-location: bottom
    keep-tex: true
    reference-doc: docs/jfwm_template.docx
editor_options:
  markdown:
    wrap: 80
execute: 
  echo: false
  warning: false
  error: false
  freeze: auto
---

# Abstract

Seagrass restoration increasingly relies on seed-based approaches, yet optimal
seeding densities and deployment methods remain poorly resolved. We evaluated
how seed density and clam-mediated seed placement influence eelgrass (*Zostera
marina*) establishment in a field restoration experiment in Back Sound, North
Carolina. Eelgrass seeds were attached to juvenile live clams at densities
ranging from 10 to 100 seeds per clam, and deployed in a fully factorial design
crossing seed density with two clam placement treatments - manually buried or
placed on the sediment surface. Plots were monitored biweekly for shoot
establishment and growth over six months. Shoot density increased through time
and was generally higher in plots receiving greater numbers of seeds. However,
the proportion of seeds that produced shoots did not vary with seed density,
indicating that higher seeding rates yielded more shoots but did not increase
per-seed establishment success. We found no evidence that clam placement
influenced seedling establishment. Similarly, shoot growth trajectories were
broadly comparable between clam placement treatments, though higher seed
densities generally yielded taller shoots. Extensive clam predation and
disturbance occurred shortly after deployment, potentially contributing to
unexplained variation among plots and influencing treatment effects. Our results
suggest that increasing seed inputs can increase restoration outcomes by
producing more established shoots, but does not improve the probability that
individual seeds establish. These findings suggest that seed density per
deployment unit can be selected based on desired restoration configuration and
design. Furthermore, manually burying clams provided little measurable benefit
over surface placement, suggesting that clams likely sufficiently bury seeds
through natural behavior and therefore this less labor-intensive deployment
method may be sufficient for large-scale restoration efforts.

# Introduction

Seagrass meadows provide important ecosystem services, including habitat for
marine organisms, shoreline stabilization, and carbon storage, but they continue
to decline in many regions worldwide. As a result, restoration has become an
increasingly important tool for recovering lost seagrass habitat. Seed-based
restoration is especially attractive because it can be scaled to larger areas
than traditional transplant-based approaches, but establishment rates are often
low and highly variable.

One of the major challenges in seed-based restoration is determining how many
seeds should be deployed. Increasing seed density may increase the total number
of seedlings that establish simply because more seeds are available. However,
establishment may not increase proportionally if seeds or seedlings begin
competing with one another or if other factors limit success. From a restoration
perspective, it is important to know whether adding more seeds increases the
number of established shoots, improves the probability that individual seeds
establish, or both. It is also unclear whether increasing seed density results
in proportional increases in shoot establishment or whether density-dependence
processes alter establishment success.

Recent restoration efforts have also explored using clams as a seed delivery
mechanism. As clams move and bury themselves in the sediment, they may help
place seeds in conditions that promote germination and reduce seed loss. There
are also positive interactions between clams and seagrass that can improve
establishment success, potentially through nutrient enhancement (Zhang).
However, manually burying clams requires additional effort and may not be
practical at large scales, though clams on the surface remain vulnerable to
predation. Whether manual burial is necessary to achieve restoration outcomes
remains unclear.

In this study, we tested how seed density and clam placement influence eelgrass
(*Zostera marina*) establishment in a field restoration experiment in North
Carolina. Specifically, we asked whether increasing seed density increases
establishment, whether manually burying clams improves establishment relative to
placing them on the sediment surface, and whether these effects interact. We
quantified both total shoot establishment and per-seed establishment to evaluate
how restoration efficiency changes across a range of seeding densities.

# Methods

## *Seed retrieval and storage*

Reproductive *Zostera marina* shoots were collected from seagrass beds in Back
Sound, NC in May 2025. Shoots were held in flow-through seawater tanks indoors
until mature seeds were naturally realeased. Released seeds were collected from
the bottom of the tank and stored in mesh bags in a flow-through tank. While
this storage tank was exposed to general ambient light conditions, the mesh bags
reduced overall light exposure. Before deployment, 100 seeds were randomly
selected to test for viability using tetrazolium staining (REF), with 89% of
seeds classified as viable.

## *Planting experiment*

To test how seed planting density and clam placement influences seagrass
establishment and growth, we conducted a field experiment near Harker’s Island,
North Carolina. Seagrass seeds were attached to live clams (*Mercenaria
mercenaria;* \~2 cm shell length) using a gelatin-based bioglue at one of ten
density levels, spanning 10-100 seeds per clam in increments of 10 seeds. Seed
planting density was crossed with two clam placement treatments (manually buried
or placed on sediment surface) in a fully factorial design with four replicates
per treatment combination (80 plots total).

The bioglue was prepared by dissolving dry gelatin powder in saltwater (35 ppt)
to a final concentration of 9% (w/v). The solution was heated to 35°C until
fully dissolved and then cooled to 28°C before application. A thin layer of
bioglue was pipetted onto one shell from each clam and allowed to partially set
before seeds were arranged in a single layer and covered with a second thin coat
of glue. Clams were refrigerated at 9°C for 30 minutes to allow the glue to
solidify, then transported to the field in coolers and deployed the same day.
The total time from initial glue preparation through field deployment was less
than four hours.

Clams were either manually buried to a depth of 2 cm or placed on the sediment
surface to allow natural self-burial. Each clam was positioned at the center of
a 0.25 m^2^ plot, with the corner of each plot marked with a lawn staple to
facilitate resampling. Plots were arranged in a grid with 0.5 m spacing in a
sandy area located at least 10 m away from adjacent salt marsh, oyster reef, and
seagrass habitat.

Plots were monitored approximately every two weeks from November 2025 through
April 2026. During each survey, we recorded the number of live shoots and
measured the height of up to five randomly selected shoots to the nearest
millimeter. During the first four sampling dates, we assessed clam presence
using brief visual and tactile confirmation. We avoided excavating clams to
limit seedling disturbance. We recorded evidence of clam predation (e.g., shell
fragments, empty shells), and whether the clam was live, dead, or not located.

```{r setup}
# Setup -------------------------------------------------------------------------
library(readxl)
library(tidyverse)
library(lme4)
library(lmerTest)
library(glmmTMB)
library(DHARMa)
library(gt)
library(ggeffects)
library(emmeans) 
library(patchwork)
library(broom.mixed)

# Helper function that will add significance to p values:
add_significance <- function(df) {df %>%
    mutate(significance = factor(case_when(p_value < 0.001 ~ "***",
                                           p_value < 0.01 ~ "**",
                                           p_value < 0.05 ~ "*",
                                           is.na(p_value) ~ "NA",
                                           TRUE ~ "NS"), 
                                 levels = c("***", "**", "*", "NS", "NA")))
}

# Read data from Excel file 
dat_raw <- read_xlsx("data/seed-density.xlsx", na = "NA") 

# Build ------------------------------------------------------------------------
# - density is continuous
# - burial is a factor
# - plot_id identifies the repeated-measures unit
# - time is measured as days since the first sampling date
# - shoot_count is the number of shoots observed at each sampling event

dat <- dat_raw %>% 
  mutate(plot_id = factor(plot_id),
         burial = factor(burial_treatment,
                         levels = c(0, 1),
                         labels = c("Buried", "Surface")),
         date = lubridate::ymd(date),
         time_days = as.numeric(date - date('2025-10-31')),
         length_cm_5 = as.numeric(length_cm_5)) %>% 
  # monitoring on 4 Jan 2026 was incomplete because poor conditions (remove)
  filter(date != '2026-01-04') %>% 
  # center density once throughout
  mutate(density_c = density - mean(density, na.rm = T),
         # quadratic term in case we want it for modeling
         density_c2 = density_c^2) %>% 
  # bin densities in case useful for later plotting
  mutate(density_group = case_when(density <= 30 ~ "Low\n10–30 seeds",
                                   density <= 70 ~ "Medium\n40–70 seeds",
                                   density >= 80 ~ "High\n80–100 seeds"),
         density_group = factor(density_group,
                                levels = c("Low\n10–30 seeds", "Medium\n40–70 seeds",
                                           "High\n80–100 seeds"))) %>% 
  # calculate proportion established relative to seed density
   mutate(est_prob = shoot_count/density) %>% 
   mutate(failures = density - shoot_count)

  
```

## *Data analyses*

To evaluate how seed planting density and clam placement influenced restoration
outcomes, we quantified three complementary response variables: shoot density,
per-seed establishment, and shoot growth. Shoot density was quantified as the
number of shoots per plot (0.0625 m^2^). Per-seed establishment was quantified
as the number of observed live shoots relative to the number of seeds initially
planted in each plot. Both shoot density and per-seed establishment were
analyzed for the final sampling date: `r format(max(dat$date), "%B %d, %Y")`.
Shoot growth was quantified as mean shoot height through time.

Final shoot density was analyzed using a generalized linear model with a
negative binomial distribution and log link. Per-seed establishment was analyzed
using a beta-binomial distribution with a logit link. Both models included seed
planting density (continuous), clam placement (factor; buried or surface), and
their interaction as fixed effects. Mean shoot height was analyzed using a
linear mixed-effects model with plot included as a random intercept to account
for repeated measurements through time. Fixed effects for included seed planting
density, clam placement treatment, sampling date, and the interaction between
seed planting density and sampling date.

Significance of fixed effects was evaluated using Type II ANOVA ($\chi^2$
tests), and model assumptions were assessed using simulated residual diagnostics
(DHARMa; REF) to evaluate dispersion and zero inflation. Model predictions and
estimated marginal means (emmeans; REF) were used to visualize treatment effects
and quantify effect sizes. All analyses were conducted in R (version X; REF) &
RStudio (version X; REF).

```{r build-dataframes}
#| include: false

# Calculate treatment-level means through time
dat_means <- dat %>% 
  group_by(date, density, burial) %>% 
  summarize(shoot_count_mean = mean(shoot_count),
            shoot_count_sd = sd(shoot_count),
            shoot_count_n = n()) %>% 
  mutate(shoot_count_se = shoot_count_sd/sqrt(shoot_count_n))

# Create predation dataframe from clam checks
pred <- dat %>% 
  select(date:dead_clam, burial) %>% 
  filter(!is.na(clam_code)) %>%
  filter(date != "2026-04-01") %>% 
  mutate(date = as.factor(date),
         clam_code_label = case_when(clam_code == "D" ~ "Empty shell",
                                     clam_code == "L" ~ "Live",
                                     clam_code == "S" ~ "Shell fragments",
                                     clam_code == "N" ~ "Nothing"))

# Create growth dataframe
dat_growth <- dat %>% 
  select(date, plot_id, density, burial, time_days, 
         shoot_count, starts_with("length")) %>% 
  pivot_longer(starts_with("length"), 
               names_to = "length_replicate", values_to = "length_cm") %>%
  filter(!is.na(length_cm)) %>% 
  group_by(date, plot_id, density, burial, time_days) %>% 
  summarize(length_cm_mean = mean(length_cm) ,
            length_cm_sd   = sd(length_cm),
            length_cm_n    = n(), .groups = 'drop') %>% 
  mutate(length_cm_se = length_cm_sd/sqrt(length_cm_n)) %>% 
  mutate(time_factor = factor(time_days,
                              levels = c('47', '78','91','106','152','168'),
                              labels = c("Dec 17", "Jan 17", "Jan 30", "Feb 14", "Apr 1", "Apr 17")))

# Build "shoot establishment" models using last sampling date (April)
dat_end <- dat %>% 
  filter(date == max(date))

# Summarize to show treatment means and error
dat_end_summary <- dat_end %>% 
  group_by(density, burial) %>% 
  summarize(mean_count = mean(shoot_count),
            se_count = sd(shoot_count) / sqrt(n()),
            .groups = "drop")

```

```{r m_count}
#| include: false

# Test whether more seeds = more shoots  
m_count <- glmmTMB(shoot_count ~ density * burial, family = nbinom2, data = dat_end)

summary(m_count)

# Check diagnostics
res_count <- simulateResiduals(m_count)
plot(res_count)
testDispersion(res_count)
testZeroInflation(res_count)

# Check the fixed effects and save outputs for results text
m_count_tests <-  car::Anova(m_count, type = 2) %>% 
  rownames_to_column("term") %>% 
  rename(p = `Pr(>Chisq)`) %>% 
  janitor::clean_names() %>% 
  mutate(chisq = round(chisq, 2),
         p = round(p, 3))

# Get the values for the results text
# From the m_count model:
density_chisq <- filter(m_count_tests, term == "density") %>% pull(chisq)
density_p     <- filter(m_count_tests, term == "density") %>% pull(p) %>% scales::pvalue()
burial_chisq  <- filter(m_count_tests, term == "burial")  %>% pull(chisq)
burial_p      <- filter(m_count_tests, term == "burial")  %>% pull(p) %>% scales::pvalue()
int_chisq     <- filter(m_count_tests, term == "density:burial") %>% pull(chisq)
int_p         <- filter(m_count_tests, term == "density:burial") %>% pull(p) %>% scales::pvalue()

# Get the expected shoots counts for 10 vs 100 seeds, averaged over burial treatment
m_count_pred <- emmeans(m_count, ~ density, 
                        at = list(density = c(10, 100)),
                        type = "response") %>%  
  summary() %>% 
  janitor::clean_names() %>% 
  mutate(across(c(response, se, asymp_lcl, asymp_ucl), ~round(.x, 3)))

shoot10      <- filter(m_count_pred, density == 10) %>% pull(response) %>% round(., 1)
shoot10_low  <- filter(m_count_pred, density == 10) %>% pull(asymp_lcl) %>% round(., 1)
shoot10_high <- filter(m_count_pred, density == 10) %>% pull(asymp_ucl) %>% round(., 1)

shoot100      <- filter(m_count_pred, density == 100) %>% pull(response) %>% round(., 1)
shoot100_low  <- filter(m_count_pred, density == 100) %>% pull(asymp_lcl) %>% round(., 1)
shoot100_high <- filter(m_count_pred, density == 100) %>% pull(asymp_ucl) %>% round(., 1)

# Estimate the average density effect
m_count_trend <- emtrends(m_count, ~1, var = "density") |>
  summary()

# Convert avg density effect te expected % increase in final shoot count per 10 additional seeds
pct10      <- round((exp(m_count_trend$density.trend * 10) - 1) * 100, 1)
pct10_low  <- round((exp(m_count_trend$asymp.LCL * 10) - 1) * 100, 1)
pct10_high <- round((exp(m_count_trend$asymp.UCL * 10) - 1) * 100, 1)

# Generate predictions for the figure
m_count_pred_all <- ggpredict(m_count, terms = c("density[10:100]", "burial")) %>% 
  rename(density = x, burial = group)

```

```{r m_prop}
#| include: false

# Start with a binomial mixed model where the response is the number of shoots relative to the initial seeds. Each plot started with a set number of seeds, and each sampling event we see a number of shoots. As a binomial process, the "successes" are the seeds that produce shoots and the "failures" are seeds that didn't. Answers the question "out of # seeds, how many became shoots?" It also accounts for the different starting number of seeds (different number of 'chances') and explicitly models the probability of establishment per seed, answering questions about density-dependent net shoot establishment.

# Test if higher density changes establishment efficiency
m_prop <- glmmTMB(cbind(shoot_count, density - shoot_count) ~ density * burial,
                  family = betabinomial(), data = dat_end)

summary(m_prop)

# Check diagnostics
res_prop <- simulateResiduals(m_prop)
plot(res_prop)
testDispersion(res_prop)
testZeroInflation(res_prop)

# Check the fixed effects
m_prop_tests <- car::Anova(m_prop, type = 2) %>% 
  rownames_to_column("term") %>% 
  rename(p = `Pr(>Chisq)`) %>% 
  janitor::clean_names() %>% 
  mutate(chisq = round(chisq, 2),
         p = round(p, 3))

# Get the values for the results text
# From the m_prop model:
m_prop_density_chisq <- filter(m_prop_tests, term == "density") %>% pull(chisq)
m_prop_density_p     <- filter(m_prop_tests, term == "density") %>% pull(p)
m_prop_burial_chisq  <- filter(m_prop_tests, term == "burial")  %>% pull(chisq)
m_prop_burial_p      <- filter(m_prop_tests, term == "burial")  %>% pull(p) 
m_prop_int_chisq     <- filter(m_prop_tests, term == "density:burial") %>% pull(chisq)
m_prop_int_p         <- filter(m_prop_tests, term == "density:burial") %>% pull(p) 

# Generate predictions for the figure
m_prop_pred_all <- ggpredict(m_prop, terms = c("density[10:100]", "burial")) %>% 
  rename(density = x, burial = group)

```

```{r m_growth}
#| include: false
#| 
# Build model that examines shoot height through time
m_growth <- lmer(length_cm_mean ~ density * time_factor + burial + (1 | plot_id),
               data = dat_growth)

m_growth2 <- lmer(length_cm_mean ~ density * time_days * burial + (1 | plot_id),
               data = dat_growth)

m_growth3 <- lmer(length_cm_mean ~ density * time_factor * burial + (1 | plot_id),
               data = dat_growth)

m_growth4 <- lmer(length_cm_mean ~ density * time_days + burial + (1 | plot_id),
               data = dat_growth)

anova(m_growth, m_growth2, m_growth3, m_growth4) 

# Most supported model has time as a factor and only 2-way interaction density * time

summary(m_growth)

# Check out the diagnostics 
growth_res <- simulateResiduals(m_growth)
plot(growth_res) # wow looks surprisingly good

# Quick check to see what this looks like
plot(effects::allEffects(m_growth), multiline = T, confint = list(style = 'auto'))
# seems like differences in densities emerged in Feb and continued from there

# Check fixed effects with the anova:
m_growth_tests <- car::Anova(m_growth, type = 2) %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  rename(p = `Pr(>Chisq)`) %>% 
  janitor::clean_names() %>% 
  mutate(chisq = round(chisq, 2),
         p = round(p, 3))

# Get the values for the results text
# From the m_growth model:
m_growth_density_chisq <- filter(m_growth_tests, term == "density") %>% pull(chisq)
m_growth_density_p     <- filter(m_growth_tests, term == "density") %>% pull(p)
m_growth_time_chisq    <- filter(m_growth_tests, term == "time_factor")  %>% pull(chisq)
m_growth_time_p        <- filter(m_growth_tests, term == "time_factor")  %>% pull(p) 
m_growth_burial_chisq  <- filter(m_growth_tests, term == "burial")  %>% pull(chisq)
m_growth_burial_p      <- filter(m_growth_tests, term == "burial")  %>% pull(p) 
m_growth_int_chisq     <- filter(m_growth_tests, term == "density:time_factor") %>% pull(chisq)
m_growth_int_p         <- filter(m_growth_tests, term == "density:time_factor") %>% pull(p) 


# Shoot height increased through time, and the effect of planting density 
# varied among sampling dates (density × sampling date interaction, Table X).

# Use emtrends to identify when the differences in heights across densities emerged
emtrends(m_growth, ~ time_factor, var = "density")


emm <- emmeans(m_growth, ~ density | time_factor,
               at = list(density = c(10, 100)))

pairs(emm)

# Generate predictions for the figure
growth_pred <- ggpredict(m_growth, 
                         terms = c("time_factor", "density [10,20,40,60,80,100]"))

```

# Results

Shoot establishment increased significantly with seed planting density ($\chi^2$
= `r density_chisq`, *p* = `r density_p`), whereas both clam placement ($\chi^2$
= `r burial_chisq`, *p* = `r burial_p`) and the interaction between seed
planting density and clam placement ($\chi^2$ = `r int_chisq`, *p* = `r int_p`)
were not significant (@fig-1). Each additional 10 seeds planted increased the
expected number of established shoots by `r pct10`% (95% CI:
`r pct10_low`-`r pct10_high`%). Model predictions indicated that increasing
planting density from 10 to 100 seeds increased the expected number of
established shoots from `r shoot10` (95% CI: `r shoot10_low`-`r shoot10_high`)
to `r shoot100` (95% CI: `r shoot100_low`-`r shoot100_high`).

Shoot establishment increased throughout the experiment across all treatments
(@fig-2). Following the onset of shoot establishment in December, higher
planting densities accumulated shoots more rapidly than lower planting
densities, resulting in progressively greater differences in shoot establishment
through time.

This increase reflected greater shoot establishment rather than higher per-seed
establishment: while shoot density increased with seed density, the *proportion*
of seeds that produced shoots did not. Per-seed establishment was not
significantly affected by seed density ($\chi^2$ = `r m_prop_density_chisq`*, p*
= `r m_prop_density_p`), clam placmeent ($\chi^2$= `r m_prop_burial_chisq`, *p*
= `r m_prop_burial_p`), or their interaction ($\chi^2$ = `r m_prop_int_chisq` ,
*p* = `r m_prop_int_p`) (@fig-1).

Mean shoot height increased throughout the experiment ($\chi^2$ =
`r m_growth_time_chisq` , *p =* `r m_growth_time_p`), and the relationship
between seed planting density and shoot height varied across sampling dates
(seed density x sampling date: $\chi^2$ = `r m_growth_int_chisq` , *p =*
`r m_growth_int_p`). Clam placement did not affect shoot height ($\chi^2$ =
`r m_growth_burial_chisq` , *p =* `r m_growth_burial_p`). Model predictions
indicated that shoot height was similar among planting densities during the
first three months, whereas shoots in higher-density treatments were
consistently taller during the latter half of the experiment (@fig-3).

# Discussion

-   Seed density influenced the number of established shoots, but not the
    probability of establishment

    -   Increasing seed density produced a predictable increase in final shoot
        establishment.

    -   Establishment probability remained consistent across the densities
        tested, indicating little evidence for density-dependent effects during
        early establishment.

    -   This is important because many restoration studies implicitly assume
        higher local propagule densities could either enhance establishment
        (facilitations) or reduce it (competition). We suggest neither process
        dominates over the density range tested.

-   Why?

    -   Tested range perhaps did not reach densities where either facilitation
        or competition among establishing seedlings became important.

    -   Other processes were more limiting than local density

-   Restoration design may be most effectively based on target shoot density and
    configuration rather than concerns about per-seed establishment.

    -   We manipulated seed density per deployment unit (clam). Since
        establishment probability was similar across densities, higher seed
        density per clam produced more shoots concentrated around one deployment
        location. Lower seed densities distributed across more clams would be
        expected to produce a similar total number of shoots but different
        spatial arrangement.

    -   Future work could explore optimal spacing across deployment units, patch
        expansion with different initial seed densities, and broader landscape
        configuration.

-   Growth trajectories also differed across planting densities, suggesting that
    there are some differences that extend beyond initial establishment. This
    was

# Figures

```{r m_count-figure}
#| fig-cap: "Relationship between initial seed density and shoot density at the final sampling point. Shoot density is the number of observable shoots, reflecting the combined effects of seed retention, germination, and early seedling survival. Points represent individual plots; lines show model predictions (±95% CI) by clam placement treatment, on the sediment surface (green circles, solid line) or manually buried (purple triangles, dashed line)."

plot_m_count <- ggplot() +
  geom_point(data = dat_end, 
             aes(x = density, y = shoot_count, 
                 color = burial, shape = burial),
             alpha = 0.8, size = 2.5,
             position = position_jitter(width = 1.5, height = 0)) +
  geom_ribbon(data = m_count_pred_all,
              aes(x = density, ymin = conf.low, ymax = conf.high, fill = burial),
              alpha = 0.15, color = NA) +
  geom_line(data = m_count_pred_all,
            aes(x = density, y = predicted, color = burial, linetype = burial), 
            linewidth = 1.2) +
  # geom_point(data = dat_end_summary, 
  #            aes(x = density, y = mean_count, 
  #                fill = burial, color = burial, shape = burial), size = 3,
  #            position = position_dodge(width = 3)) +
  # geom_errorbar(data = dat_end_summary, 
  #               aes(x = density, ymin = mean_count - se_count, ymax = mean_count + se_count,
  #                   color = burial),
  #               width = 2,linewidth = 0.5,
  #               position = position_dodge(width = 3)) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 6) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  scale_fill_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = "Seed planting density",
       y = "Number of shoots per plot",
       linetype = "Clam placement",
       color = "Clam placement",
       fill = "Clam placement",
       pch = "Clam placement") +
  theme_classic() +
  theme(legend.position = c("inside"),
        legend.position.inside = c(0.2, 0.85))

```

```{r m_prop-figure}
#| fig-cap: "Relationship between initial seed density and shoot establishment at the final sampling point. Establishment is defined as the proportion of seeds that resulted in observable shoots, reflecting the combined effects of seed retention, germination, and early seedling survival. Points represent individual plots; lines show model predictions (±95% CI) by clam placement treatment, on the sediment surface (green circles, solid lie) or manually buried (purple triangles, dashed line)."
# Mean shoot height increased throughout the experiment, and the relationship between 
# planting density and shoot height varied across sampling dates 
# (density × sampling date interaction; Table X). 
# Although the estimated effect of planting density was weakest early in the experiment
# and strongest during the latter sampling periods, differences among sampling dates 
# were not statistically distinguishable after adjustment for multiple comparisons.

plot_m_prop <- ggplot() +
  geom_point(data = dat_end,
             aes(x = density, y = est_prob, color = burial, fill = burial, shape = burial),
             alpha = 0.8, size = 2.5,
             position = position_jitter(width = 1.5, height = 0)) +
  geom_ribbon(data = m_prop_pred_all,
              aes(x = density, ymin = conf.low, ymax = conf.high, fill = burial),
              alpha = 0.15, color = NA) +
  geom_line(data = m_prop_pred_all,
            aes(x = density, y = predicted, color = burial, linetype = burial), 
            linewidth = 1.2) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 6) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  scale_fill_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = "Seed planting density",
       y = "Proportion of seeds producing shoots",
       linetype = "Clam placement",
       color = "Clam placement",
       fill = "Clam placement",
       pch = "Clam placement") +
  theme_classic() +
  theme(legend.position = c("inside"),
        legend.position.inside = c(0.9, 0.8))

```

```{r fig-1}
#| label: fig-1
#| fig-width: 4
#| fig-height: 7
#| out-width: "3.5in"
#| fig-dpi: 300
#| fig-cap: "Effects of seed planting density on (A) final shoot count and (B) establishment probability at the final sampling point. Establishment probability was calculated as the proportion of planted seeds that produced observable shoots, integrating seed retention, germination, and early seedling survival. Points represent individual plots. Lines show model predictions (±95% CI) by clam placement treatment: on the sediment surface (purple triangles, dashed line) or manually buried (green circles, solid line)."

plot_m_prop  <- plot_m_prop  + theme(legend.position = "none")

plot_seed_outcomes <- (plot_m_count / plot_m_prop) +
  plot_annotation(tag_levels = "A")

plot_seed_outcomes
```

```{r fig-2}
#| label: fig-2
#| fig-width: 6
#| fig-height: 4.5
#| out-width: "6in"
#| fig-dpi: 300
#| fig-cap: "Temporal patterns in shoot establishment across seed planting density groups and clam placement treatments. For visualization, seed planting densities were grouped into low (10-30 seeds), medium (40-70 seeds), and high (80-100 seeds) categories. Points represent the average shoot count per plot (±SE) at each sampling date, averaged across all plots within each seed density group and clam placement treatment. Green circles and solid lines indicate manually buried clams; purple triangles and dashed lines indicate clams placed on the sediment surface."

dat_density_group_means <- dat %>%
  group_by(date, burial, density_group) %>%
  summarise(shoot_count_mean = mean(shoot_count, na.rm = TRUE),
            shoot_count_se = sd(shoot_count, na.rm = TRUE) / sqrt(n()),
            establishment_mean = mean(shoot_count / density, na.rm = TRUE),
            establishment_se = sd(shoot_count / density, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

ggplot(dat_density_group_means,
       aes(date, shoot_count_mean, 
           color = burial, fill = burial, group = burial, pch = burial, linetype = burial)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = shoot_count_mean - shoot_count_se,
                    ymax = shoot_count_mean + shoot_count_se),
                alpha = 0.2, color = NULL) +
  facet_wrap(~density_group) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  scale_fill_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = "",
       y = expression("Shoot count per 0.0625 m"^2),
       color = "Clam placement",
       fill = "Clam placement",
       pch = "Clam placement",
       linetype = "Clam placement") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

```

```{r fig-3}
#| label: fig-3
#| fig-width: 4.5
#| fig-height: 4
#| out-width: "4.5in"
#| fig-dpi: 300
#| fig-cap: "Effects of seed planting density on mean shoot height (cm) over the course of the experiment. Lines show model-predicted mean shoot height (±95% CI) from a linear mixed effects model, averaged across clam placement treatments. Darker blue lines indicated higher seed densities."

density_cols <- c("#8CB9D8", "#6E9FC8", "#4F86B7", "#356DA5", "#245591", "#163D73")

ggplot(growth_pred, aes(x = x, y = predicted, color = group, group = group)) +
  geom_line(linewidth = 0.9) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group),
              alpha = 0.12,
              color = NA) +
  scale_color_manual(values = density_cols) +
  scale_fill_manual(values = density_cols) +
  labs(x = "",
       y = "Mean shoot height (cm)",
       color = "Seed density",
       fill = "Seed density") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

```

Other potential figures:

-   Early predation effects on establishment, with seed density on x-axis and
    establishment probability on the y-axis. Panels for whether or not there was
    evidence of predation?

-   Alternative could make boxplots with final time point based on predation.
    Explore whether earlier disturbance suppresses establishment (and maybe
    explain null treatment effects? or just context for the predation
    observations).

-   Heatmap showing clam presence over time for first few sampling dates...
    illustrate level of predation? Also could illustrate whether burial reduces
    predation risk (but I don't think so).

Fig 1: No strong density or burial effect\
Fig 2: Time dominates establishment\
Fig 3: Growth also largely insensitive\
Fig 4: Early predation explains variation

“We expected density and burial to matter; they did not strongly. Instead, early
disturbance and temporal dynamics structure establishment.”

# **Notes**

[*Storage copied from Mason's MS:*]{.underline}

Storage tanks were housed in an indoor facility equipped with bubblers
and maintained on a 12-hour light cycle. Tanks were stirred daily to prevent
shoot desiccation and to encourage seed release. After 4 weeks, excess plant
material was removed by hand. Released seeds were panned to separate them from
extraneous material and maintained in flow-through tanks until October 2025. 

Juvenile clams were sourced from Carteret Community College located in Morehead
City, North Carolina and maintained in a flow-through seawater system at the
Duke University Marine Laboratory prior to planting. 

Planting timing coincided with observed natural germination periods in North
Carolina, when water temperatures consistently reach 15 °C or below,
the optimal temperature for Zostera marina seed germination. 

[*Deployed Friday October 31, 2025 at 12pm (during low tide).*]{.underline}

We marked the corners of each plot with tall bamboo poles (3 m) hammered into
the sediment using a rubber mallet while standing on the stern of the boat. We
established transects for each row in the grid, marking the start and end of
each transect with small bamboo poles (1 m) sunk into the sediment (left
permanently during the experiment to facilitate resampling). Rows were oriented
roughly parallel to the adjacent marsh shoreline, with seed density treatments
increasing each row from the side of the site closest to the marsh edge (lowest
density) to the outermost row (highest density). Note that the site itself is
\~50m from the marsh edge. Facing the marsh, the four columns on the left side
of the site contained manually buried clams, and the four columns on the right
side of the site contained clams placed on the sediment surface (see site
diagram, below).

On at least two occasions, the bioglue detached from the clam during underwater
handling, though the seeds remained intact within the glue matrix. These cases
were treated as valid experimental conditions rather than handling errors, and
both the detached bioglue-seed cluster and the corresponding clam were placed in
the assigned treatment position. Similar detachment events were anecdotally
observed during other planting experiments that season. 

Conditions during deployment were fairly windy (15-25 mph) with air temperatures
18.3°C and water temperatures 21.7°C. Visibility was poor; less than 0.5 m,
which prevented any visual confirmation of activity of “surface” clams (e.g.,
movement, burial). Deployment was \~30 minutes after low tide (+0.47ft) and
water depth was \~0.5-0.8 m during deployment (rising; knee to hip). Sediment
was sand covered by a thin layer of mud with minimal detritus, though some
sparse *Halodule wrightii* was observed at the margins of the site, especially
towards the outermost row (highest density treatments). Upon re-examining the
site under better conditions, a few plots contained sparse *Halodule* within the
plot.

[*Site diagram (for internal purposes)*]{.underline}

*(All plots 0.25m2 and are equally spaced by 0.5m2)*

Letters mark tall bamboo poles, and x's mark short bamboo poles used to lay out
rows. Pole B is closest to Harker’s Island Boat Ramp. A-B parallel to marsh, B-C
closest to bridge and parallel to oysters, A-D farthest to bridge parallel
opposite oysters, D-C closest to entrance/exit to study area and adjacent
seagrass beds.

\~\~\~\~ MARSH EDGE, Mason + Christy Experiments \~\~\~\~

| A     |      |      |      |      |      |      |      |      | B     |
|-------|------|------|------|------|------|------|------|------|-------|
| x     | B10  | B10  | B10  | B10  | S10  | S10  | S10  | S10  | x     |
| x     | B20  | B20  | B20  | B20  | S20  | S20  | S20  | S20  | x     |
| x     | B30  | B20  | B20  | B20  | S30  | S30  | S30  | S30  | x     |
| x     | B40  | B40  | B40  | B40  | S40  | S40  | S40  | S40  | x     |
| x     | B50  | B50  | B50  | B50  | S50  | S50  | S50  | S50  | x     |
| x     | B60  | B60  | B60  | B60  | S60  | S60  | S60  | S60  | x     |
| x     | B70  | B70  | B70  | B70  | S70  | S70  | S70  | S70  | x     |
| x     | B80  | B80  | B80  | B80  | S80  | S80  | S80  | S80  | x     |
| x     | B90  | B90  | B90  | B90  | S90  | S90  | S90  | S90  | x     |
| x     | B100 | B100 | B100 | B100 | S100 | S100 | S100 | S100 | x     |
| **D** |      |      |      |      |      |      |      |      | **C** |

\~\~\~\~ ACCESS EDGE, towards deeper water \~\~\~\~

*Descriptive visualizations*

We conducted initial exploratory analyses to characterize temporal patterns in
shoot density across seed density and clam placement treatments. First, we
visualized plot-level trajectories of shoot densities through time, stratified
by seed density to explore whether temporal dynamics differed across density
treatments and whether patterns varied with clam placement.

To highlight treatment-level patterns, we calculated average shoot density
across replicates for each sampling date, seed density, and clam placement
treatment (n = 4 plots for each combination). We plotted the temporal
trajectories of these treatment-level means to examine general trends in shoot
density over time and explore potential differences between clam placement
treatments. These descriptive analyses were used to identify broad patterns (or
lack thereof) and inform model structure.

As shoot density is inherently constrained by the number of seeds planted, we
also examined shoot density relative to planted seed density. Probability of
shoot establishment was quantified as the number of live shoots relative to the
number of seeds initially planted in each plot.

[Trajectories of raw shoot densities]{.underline}

Initial visualizations show that shoot densities increase over time, and that
higher seed densities generally yield higher shoot densities (Figure 1). This
pattern does not appear to vary depending on whether clams were manually buried
or placed on the sediment surface. For visualization purposes, these trends are
shown for each plot through time (Figure 1) as treatment-level means through
time (Figure 2), and as treatment-level means through time but grouped by clam
placement (Figure 3).

```{r plot-level-density-trajectories}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of shoot density (count per 0.0625 m²) from November 2025 through April 2026 across experimental plots, shown separately for initial seed densities from 10 to 100 seeds (facets). Each line represents a single plot. Clam placement treatments are indicated by symbol and line type: clams placed on the sediment surface are shown as green circles with solid lines, and manually buried clams as purple triangles with dashed lines."

ggplot(dat, aes(x = date, y = shoot_count, group = plot_id, color = burial)) +
  geom_line(aes(linetype = burial), alpha = 0.7) +
  geom_point(aes(shape = burial), alpha = 0.7) +
  facet_wrap(~density, nrow = 2) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       shape = "Clam placement",
       linetype = "Clam placement",
       color = "Clam placement") +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r treatment-level-density-trajectories}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of mean shoot density (count per 0.0625 m²) from November 2025 through April 2026, shown separately for initial seeding densities from 10 to 100 seeds (facets). Values represent treatment-level means (n = 4 plots per density × clam placement combination) and error bars show +/- 1 standard error. Clam placement treatments are indicated by symbol and line type: clams placed on the sediment surface are shown as green circles with solid lines, and manually buried clams as purple triangles with dashed lines."

ggplot(dat_means, aes(x = date, y = shoot_count_mean, group = burial, color = burial)) +
  geom_line(aes(linetype = burial)) +
  geom_point(aes(shape = burial)) +
  geom_errorbar(aes(ymin = shoot_count_mean - shoot_count_se, 
                    ymax = shoot_count_mean + shoot_count_se)) +
  facet_wrap(~density, nrow = 2) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       shape = "Clam placement",
       linetype = "Clam placement",
       color = "Clam placement") +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(dat_means,
       aes(date, shoot_count_mean,
           color = factor(density),
           group = density)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~burial) +
  theme_classic()

dat_density_group_means <- dat %>%
  group_by(date, burial, density_group) %>%
  summarise(shoot_count_mean = mean(shoot_count, na.rm = TRUE),
            shoot_count_se = sd(shoot_count, na.rm = TRUE) / sqrt(n()),
            establishment_mean = mean(shoot_count / density, na.rm = TRUE),
            establishment_se = sd(shoot_count / density, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

ggplot(dat_density_group_means,
       aes(date, shoot_count_mean, 
           color = density_group, group = density_group)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = shoot_count_mean - shoot_count_se,
                    ymax = shoot_count_mean + shoot_count_se),
                width = 4, linewidth = 0.4) +
  facet_wrap(~burial) +
  labs(x = "",
       y = expression("Shoot count per 0.0625 m"^2),
       color = "Seed planting density") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r treatment-trajectories-by-placement}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Temporal trajectories of mean shoot density (count per 0.0625 m²) from November 2025 through April 2026, shown separately by clam placement treatment (facets). Values represent treatment-level means (n = 4 plots per density × clam placement combination). Colors denote initial seed density from 10 to 100 seeds, with darker shades indicating higher densities."

ggplot(dat_means %>% 
         mutate(density = factor(density)),
       aes(x = date, y = shoot_count_mean, group = density, color = density)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = shoot_count_mean - shoot_count_se, ymax = shoot_count_mean + shoot_count_se), width = 3)+
  facet_wrap(~burial, nrow = 1) +
  scale_color_manual(values = c('#d9f0ea', '#b2e2d8', '#80cdc1', '#4db6ac', '#2ca25f',
                                '#238b45','#006d2c', '#005824','#00441b','#002e12')) +
  labs(x = "",
       y = expression("Shoot density (count per 0.0625 m"^2*")"),
       color = "Seed density") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

[Shoot growth]{.underline}

```{r plot-shoot-height}
#| fig-width: 6.5
#| fig-height: 4
#| fig-dpi: 300
#| fig-cap: "Trajectories of mean shoot height (cm) from December 2025 to April 2026. Points and lines represent treatment-level means, for each combination of seed density (shades of green) and clam placement treatment (facets)."

ggplot(dat_growth %>% 
         mutate(density = factor(density)),
       aes(x = date, y = length_cm_mean, group = density, color = density)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = length_cm_mean - length_cm_se, 
                    ymax = length_cm_mean + length_cm_se), width = 3) + 
  facet_wrap(~burial, nrow = 1) +
  scale_color_manual(values = c('#d9f0ea', '#b2e2d8', '#80cdc1', '#4db6ac', '#2ca25f',
                                '#238b45','#006d2c', '#005824','#00441b','#002e12')) +
  labs(x = "",
       y = "Shoot height (cm)",
       color = "Seed density") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r pred-heatmap}
#| fig-width: 4
#| fig-height: 6
#| fig-dpi: 300

ggplot(data = pred) +
  geom_tile(aes(x = date, y = plot_id, fill = clam_code_label)) +
  scale_fill_manual(values = c("red3","green4", "grey", "orange2")) +
  labs(x = "", y = "Plot", fill = NULL) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7),
        legend.text = element_text(size = 7)) +
 ggh4x::facet_nested(density~burial, scales = "free", space = "free", independent = "y")

```
